---
title: Exercise 7<br> Analysis of stratified data
author: Centre for Research into Ecological and Environmental Modelling <br> **University of St Andrews**
date: Introduction to distance sampling<br> April 2022
output: 
  rmdformats::readthedown:
    highlight: tango
---

<div class="alert  alert-success">
  <strong>Solutions</strong> Analysis of stratified data
</div>

Reading in the data:

```{r, message=FALSE}
library(Distance)
library(kableExtra)
# Load data
data(minke)
head(minke, n=3)
# Specify truncation distance
minke.trunc <- 1.5
```

# Strata treated distinctly

Fit detection function and encounter rate separately in each strata. 

```{r, echo=T, eval=T, message=FALSE}
## Fit to each region separately - full geographical stratification
# Create data set for South
minke.S <- minke[minke$Region.Label=="South", ]
minke.df.S.strat <- ds(minke.S, truncation=minke.trunc, key="hr", adjustment=NULL)
summary(minke.df.S.strat)
# Combine selection and detection function fitting for North
minke.df.N.strat <- ds(minke[minke$Region.Label=="North", ],
                       truncation=minke.trunc, key="hr", adjustment=NULL)
summary(minke.df.N.strat)
```

# Detections combined across strata

Next we fitted a pooled detection function. 

```{r, echo=T, eval=T, message=FALSE}
minke.df.all <- ds(minke, truncation=minke.trunc, key="hr", adjustment=NULL)
summary(minke.df.all)
```

```{r, echo=F, eval=T}
# Save AIC values
aic.all <- summary(minke.df.all$ddf)$aic
aic.S <- summary(minke.df.S.strat$ddf)$aic
aic.N <- summary(minke.df.N.strat$ddf)$aic
aic.SN <- aic.S + aic.N

```

# Determining the correct stratification to use

The AIC value for the detection function in the South was `r format(aic.S, digits=4)` and the AIC for the North was `r format(aic.N, digits=4)`. This gives a total AIC of `r format(aic.SN, digits=4)`. The AIC value for the pooled detection function was `r format(aic.all, digits=4)`. Because `r format(aic.all, digits=4)` is greater than `r format(aic.SN, digits=4)`, estimation of separate detection functions in each stratum is preferable. 

```{r, echo=F}
par(mfrow=c(2,2))
plot(minke.df.S.strat, main="South")
plot(minke.df.N.strat, main="North")
plot(minke.df.all, main="All")
```

# Differing abundance estimates from stratification decision

```{r, echo=F}
# Harvest abundance estimates
est.full <- data.frame(Label=c("North","South","Total"),Estimate=rep(NA,3))
est.full[1,2] <- minke.df.N.strat$dht$individuals$N$Estimate
est.full[2,2] <- minke.df.S.strat$dht$individuals$N$Estimate
est.full[3,2] <- est.full[1,2] + est.full[2,2]

est.dfpool <- minke.df.all$dht$individuals$N[ ,c(1,2)]
```

In the full geographical stratification, both encounter rate and detection function were estimated separately for each region (or strata). This resulted in the following abundances:

```{r, echo=FALSE}
knitr::kable(est.full, caption="Abundance estimates using full geographical stratification.", digits=0) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

Next, the distances were combined to fit a pooled detection function but encounter rate was obtained for each region. This resulted in the following abundances:

```{r, echo=FALSE}
knitr::kable(est.dfpool, caption="Abundance estimates calculating encounter rate by strata and a pooled detection function.", digits=0) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

# Another approach to stratification (advanced)

An equivalent result for full geographic stratification could be produced using the `dht2` function, which does not require the disaggregation of the data set into two data sets.

```{r, echo=T, eval=T, message=FALSE}
# Geographical stratification with stratum-specific detection function 
strat.specific.detfn <- ds(data=minke, truncation=minke.trunc, key="hr", 
                           adjustment=NULL, formula=~Region.Label)
abund.by.strata <- dht2(ddf=strat.specific.detfn, flatfile=minke, 
                        strat_formula=~Region.Label, stratification="geographical")
print(abund.by.strata, report="abundance")
```

I won't say anything just now about the wrinkle I introduced with the `formula` argument in the call to `ds()`.  Recognise there is an alternative (easier) way to perform the full geographic stratification analysis without tearing apart the data.  The abundance estimates presented in the last output do not identically match the estimates shown earlier for full geographic stratification, but they are close.  The added benefit of this latter analysis is that the uncertainty in the total population size is computed within `dht2` rather than needing to be calculated manually using the delta method.

<div class="alert  alert-info">
  <strong>An aside</strong> 
  <p>
If geographic stratification were ignored, the abundance estimate of would be 18,293 minkes. This estimate is substantially larger than the estimates above. The reason is that the survey design was geographically stratified with a smaller proportion of the north stratum receiving sampling effort and a greater proportion of the southern stratum receiving survey effort. Ignoring this inequity in the unstratified analysis would lead us to believe that the more heavily sampled southern stratum is indicative of whale density throughout the study area.
</div>